\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-02 paper/paper.tex
paper updates

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world left the starting line en route to their
destination 42.195 kilometers away. Two hours, twenty-four minutes,
and four seconds later, Jemima Sumgong of Kenya would be the first to
cross the finishline and take home gold; Sumgong was just 3.5 minutes
behind her prior personal best time in the marathon. Approximately 21
minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''
\href{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}
The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit contrived
(smiling like ``Honigkuchenpferde''---a Honigkuchenpferd is a German
cookie in the shape of a horse), and the sports director of the German
Athletics Federation, Thomas Kurschilgen, stirred up controversy when
he suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins ran in the Rio marathon at least 18 minutes slower their
personal best times prior to the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, Kurschilgen's accusations were denied by the Hahner
twins, who claimed that their simultaneous finish was simply an
unintended coincidence.

@@

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Of these ten pairs,
differences in finishing times in seconds are as follows: 36, 93, 172,
319, 379, 459, 552, 671, 675, 739.  In other words, of all pairs of
runners in the Rio marathon who had a personal best difference that
was equivalent to the Hahner twins' difference, the twins had the
greatest compression based on finishing time.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For the 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiff} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we have used earlier.

Consider first the Hahner twins.  They are effectively located on the
horizontal axis (because their difference in finishing times is one
second).  However, there are many points about the Hahner's orange
dot, and this shows that, conditional on a 115 second difference in
personal best times,

In the
context of the marathon, that is, did Anna and Lisa Hahner run with an
eye on each other?

qq

As a brief overview of the race, Figure \ref{fig:secondsbehind}
describes each runner's status at the various split times on the
marathon course.  Each dot in the figure depicts a recorded split and
the number of seconds each runner was behind the race leader at the
time.  There are more dots at earlier splits due to the accumulation
of DNFs.  Regarding twins and triplets, the Estonian Luik triplets are
denoted with blue dots; there are three blue dots through the 20
kilometer split, but the third triplet did not record a half-marathon
split.  The orange dots in the figure represent the Hahner twins; as
noted in the figure, Lisa Hahner was ahead of her sister through 15
kilometers at which point Anna surged ahead.  Lastly, Figure
\ref{fig:secondsbehind} contains two red dots representing the North
Korean Kim twins; this is not visually apparent because the Kim twins
had identical split times during the marathon.

@@

Figure \ref{fig:secondsbehind} does not directly shed any light on our
motivating research question, but it does show that the Hahner twins
finished amidst many other runners.  We will come back to this point
later.

@@

To test the assertion that the Hahner twins paced themselves to finish
at the same time and with back-to-back placements, we need to compute
the probability that such a result would have occurred
unintentionally. Hence we need to know the probability distribution of
Anna and Lisa's finishing times if their runs had been independent of
each other. The challenge, of course, is that this distribution is
unknown.  % Therefore it needs to be estimated.

@@

%% THIS SECTION IS A MESS.  I JUST WANTED TO PUT SOMETHING IN WRITING.

%% Ok!  Not sure what all the notation means.

% We know that Anna and Lisa Hahner finished the marathon in Rio
% simultaneously. We don't know, however, if the finish was intentional;

The Hahner twins either acted independently $I$ or coordinated their
times $C$; these are the two possible states of the race
$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
that they ran independently, as they say they did. Hence, we are
looking to determine, %% y_a versus Y_a.  Former realized?

@@

\begin{enumerate}
\item Removing the two sets of twins and triplets from the sample,
  estimate a linear model that predicts a runner's final time ($Y_i$)
  based on her pre-Olympic personal best time ($X_i$).
\item Extract the coefficients and the variance-covariance matrix from
  this model.
\item For each runner, draw $\beta_0$ and $\beta_1$ from a multivariate normal
distribution with mean equal to the coefficients and variance equal to the
variance-covariance matrix.
\item For each runner, draw an error term ($e$) from a normal distribution with
mean equal to zero and a standard deviation equal to the standard
deviation of the model's residuals ($\sigma$).
\item For each runner, predict the final result by combining the randomly
  generated beta coefficients and error terms, $\hat{y} = \beta_0 + \beta_1 + e$.
\item Eliminate each runner from the race with a probability equal to the
percent of the total runners who did not finish in the actual race.  This
account for the likelihood of a DNF.
\item Calculate the difference in time between Anna Hahner and Lisa Hahner.
\item Calculate the difference in ranking between Anna Hahner and Lisa Hahner.
\item Repeat steps 3 through 8 ten thousand times.
\item Plot a histogram of the results in Figure \ref{fig:simdiff}
\end{enumerate}

@@

>>>>

On August 14, 2016, at 9:30 in the morning, the Women's Olympic
marathon kicked off in Rio de Janeiro, Brazil, when 156 runners from
80 countries across the world left the starting line en route to their
destination 42.195 kilometers away. Two hours, twenty-four minutes,
and four seconds later, Jemima Sumgong of Kenya would be the first to
cross the finishline and take home gold; Sumgong was just 3.5 minutes
behind her prior personal best time in the marathon. Approximately 21
minutes later, twin marathoners from Germany, Anna and Lisa Hahner,
would cross the finishline together, holding hands and celebrating a
personal victory. Although the Hahners would finish 81st and 82nd,
respectively, with times slightly more than 18 minutes slower than
corresponding personal bests, Anna Hahner would describe their joint
finish as a ``magical moment.''

The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. Many
believed the moment was a reflection of the Olympic spirit.

%% Honigkuchenpferde reference
%%
%% https://www.welt.de/sport/olympia/article157669264/Das-falsche-Laecheln-der-deutschen-Lauf-Zwillinge.html

Not everyone agreed with this rosy interpretation. The twins' happy
facial expressions at the finish were portrayed as a bit
contrived---smiling like ``Honigkuchenpferde,'' cookies in the shape
of a horse---and the sports director of the German Athletics
Federation, Thomas Kurschilgen, stirred up controversy when he
suggested that the Hahners' photo-finish was no coincidence.
Kurschilgen averred that the twins slowed down so as to finish
simultaneously and create a spectacle which would ``generate media
attention.'' Kurschilgen justified his charge with the fact that the
twins ran in the Rio marathon at least 18 minutes slower their
personal best times prior to the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, Kurschilgen's accusations were denied by the Hahner
twins, who claimed that their simultaneous finish was simply an
unintended coincidence.

@@

Of the 133 marathon finishers, there are $\binom{133}{2} = 8,778$
pairs of runners.  Of these and ignoring the Hahner twins, ten had
exactly a 115 second gap in personal best times.  Of these ten pairs,
differences in finishing times in seconds are as follows: 36, 93, 172,
319, 379, 459, 552, 671, 675, and 739.  In other words, of all pairs
of runners in the Rio marathon who had a personal best difference that
was equivalent to the Hahner twins' difference, the twins had the
greatest compression based on finishing time.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For the 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiff} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.

Consider first the Hahner twins.  They are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points about the Hahner's orange dot,
and this shows that, conditional on a 115 second difference in
personal best times, most marathoners did not have close finishing
times like the Hahnes.  Some pairs of runners with 115 second personal
best differences had finishing time differences of 1000 seconds, i.e.,
in excess of 15 minutes.  Of course the points in Figure
\ref{fig:diffdiffscatter} are not independent, but they provide a
sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

Thinking about the accusations leveled against the Hahner twins,
Figure \ref{fig:diffdiffscatter} suggests that Anna and Lisa Hahner
did indeed run with an eye on each other.  In fact, the same can be
said of the Kim twins, who ran seemingly in lockstep throughout the
entire Rio marathon.  The North Korean twins had a personal best
difference of 53 seconds and a finishing time difference of literally
zero seconds.  Beyond these twins, there were eight pairs of runners
with a 53 second personal best difference, and resulting finishing
time differences are as follows: 9, 51, 228, 340, 352, 571, 662, and
751.  As in the Hahner case, the Kim twins compressed their finishing
times more than any other pair of runners with a similar personal best
difference.

To get a sense of the lockstep nature of the Kim twins race, Figure
\ref{fig:secondsbehind} describes each runner's status at the various
split times on the marathon course.  Each dot in the figure depicts a
recorded split and the number of seconds each runner was behind the
race leader at the time.  There are more dots at earlier splits due to
the accumulation of DNFs.  The usual color scheme applies here, and
the Estonian DNF prior to the half marathon split is evident.  The
orange dots in the figure represent the Hahner twins, and as noted in
the figure Lisa Hahner was ahead of her sister through 15 kilometers
at which point Anna surged ahead.  Figure \ref{fig:secondsbehind}
contains two red dots representing the North Korean Kim twins, but
this is not visually apparent because the Kim twins had identical
split times during the entire race marathon.

@@

@@

To test the assertion that the Hahner twins paced themselves to finish
at the same time and with back-to-back placements, we need to compute
the probability that such a result would have occurred
unintentionally. Hence we need to know the probability distribution of
Anna and Lisa's finishing times if their runs had been independent of
each other. The challenge, of course, is that this distribution is
unknown.

@@

The Hahner twins either acted independently $I$ or coordinated their
times $C$; these are the two possible states of the race
$\psi = \{I, C\}$. Given that we observed Anna and Lisa's finishing
times, $Y_A = y_A$ and $Y_L = y_L$, we want to know the probability
that they ran independently, as they say they did. Hence, we are
looking to determine,

@@

\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time ($Y_i$) based
  on her pre-Olympic personal best time ($X_i$) and her best time
  squared.
\item Extract coefficient vector and the covariance matrix from this
  model.
\item For each simulated rate, draw intercept and slope estimates
  $\beta_0$, $\beta_1$, and $\beta_2$, respectively, from a
  multivariate normal distribution with mean equal to the estimated
  coefficient vector and covariance equal to the estimated covariance
  matrix.
\item For each runner, draw an error ($e$) from a normal distribution
  with mean zero and a standard deviation equal to the standard
  deviation of the model's residuals ($\sigma$).
\item For each runner, predict the final result by combining the randomly
  generated beta coefficients and error terms, $\hat{y} = \beta_0 +
  \beta_1\,X_i + \beta_2\,X_i^2 + e$.
\item Eliminate each runner from the race with a probability equal to
  the fraction of runners who did not finish the marathon.
\item Calculate the difference in time and difference in ranking
  between Anna Hahner and Lisa Hahner, assuming both finished the
  race.
\item Repeat above steps ten thousand times.
\item Plot a histogram of the results in Figure \ref{fig:simdiff}.
\end{enumerate}

@@

\section*{Conclusion}

\newpage
\section*{Further reading}

\begin{itemize}

\item \url{http://www.telegraph.co.uk/olympics/2016/08/17/german-twins-criticised-for-finishing-olympic-marathon-fun-run-h}

\item
  \url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}

\item
  \url{https://www.welt.de/sport/olympia/article157669264/Das-falsche-Laecheln-der-deutschen-Lauf-Zwillinge.html}
\end{itemize}
